Evaluating the evidence of widespread maintenance of functional composition in vertebrate communities

### 2021-10-08 paper/manuscript.Rmd
Spruce up the estimates table.

Ecological communities are experiencing unprecedented change as a result of anthropogenic pressures such as climate change, land use change, and invasive species.
Impacts of these pressures are well documented at a global scale by an accelerating global extinction rate [@barnosky2011], and fundamental changes in some of the most well-studied systems [e.g.
coral bleaching; @sully2019].
At the local scale however, species diversity tells a different story.
Recent syntheses of local trends in biodiversity over time have found no net change in local species diversity despite ongoing turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017] and evidence of a significant shifts in community composition underlying consistent species richness [@brose2016; @gotelli2017].
While communities are clearly changing, our most common species-based approaches do not fully capture the nature of that change.

@@

In order to ensure taxonomic consistency across datasets, BioTIME species were paired with trait data based on their species identifier from the Integrated Taxonomic Information System database (retrieved 09-15-2020 from the on-line database, <https://doi.org/10.5066/F7KH0KBK>), obtained through the taxadb R package [@norman2020; @rcoreteam2021].
If more than one species in the assemblage data resolved to the same identifier, observations were considered the same species.
For trait data, traits for all species of the same identifier were averaged.
Only studies with at least 75% trait coverage were included and observations for species with no trait data were excluded.
In order to have a sufficient number of species to calculate functional diversity metrics, years with fewer than 5 species observed were also excluded.

@@

To assess functional change independent of species richness we calculating the standardized effect size (SES) for each metric from null estimates [@swenson2012].
Null model corrections allow us to assess the degree to which the observed functional diversity metric deviates from the value expected by chance in a randomly assembled community.
Null estimates were calculated for each rarefied sample by randomly sampling species from the species pool for each year and randomly assigning observed abundances to species.
Species pools included all species observed for a time series.
This process was repeated 500 times to get an estimate and standard deviation of the null expectation for the metric for each rarefaction for that time series.
We used these values to calculate SES using the following formula: $SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$.
We then calculated the median SES estimate for each metric from all the rarefaction samples for a time series.
<!-- description of edge case where community = species pool -->

@@

```{r echo=FALSE}
read_csv(here::here("figures/man_model_table.csv")) %>%
  select(metric, everything()) %>%
  mutate(metric = ifelse(metric == "Jaccard_base", "Jaccard", metric)) %>%
  arrange(metric) %>%
  kable(format = "latex", caption = "Model estimates and statistics for general trend models for species richness (S) and standardized metrics. Additional model estimates can be found in the supplement.") %>%
  kable_classic() %>% 
  kable_styling(latex_options= c("scale_down", "hold_position"))
```

@@

We also need theory that links ecological process to the kinds of changes actually happening in communities.
Do the shifts we see matter?
Are there shifts we didn't even measure that are critical for function and resilience?

>>>>

Ecological communities are experiencing unprecedented change as a result of anthropogenic pressures such as climate change, land use change, and invasive species.
Impacts of these pressures are well documented at a global scale by an accelerating global extinction rate [@barnosky2011], and fundamental changes in some of the most well-studied systems [e.g. coral bleaching; @sully2019]. At the local scale however, species diversity tells a different story. Recent syntheses of local trends in biodiversity over time have found no net change in local species diversity despite ongoing turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017] and evidence of a significant shifts in community composition underlying consistent species richness [@brose2016; @gotelli2017; @li2020].
While communities are clearly changing, our most common species-based approaches do not fully capture the nature of that change.

@@

In order to ensure taxonomic consistency across datasets, BioTIME species were paired with trait data based on their species identifier from the Integrated Taxonomic Information System database (retrieved 09-15-2020 from the on-line database, <https://doi.org/10.5066/F7KH0KBK>), obtained through the *taxadb* R package [@norman2020; @rcoreteam2021].
If more than one species in the assemblage data resolved to the same identifier, observations were considered the same species.
For trait data, traits for all species of the same identifier were averaged.
Only studies with at least 75% trait coverage were included and observations for species with no trait data were excluded.
In order to have a sufficient number of species to calculate functional diversity metrics, years with fewer than 5 species observed were also excluded.

@@

To assess functional change independent of species richness we calculating the standardized effect size (SES) for each metric from null estimates [@swenson2012]. Null model corrections allow us to assess the degree to which the observed functional diversity metric deviates from the value expected by chance in a randomly assembled community. Null estimates were calculated for each rarefied sample by randomly sampling species from the species pool for each year and randomly assigning observed abundances to species. Species pools included all species observed for a time series. This process was repeated 500 times to get an estimate and standard deviation of the null expectation for the metric for each rarefaction for that time series. We used these values to calculate SES using the following formula: $SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$.
We then calculated the median SES estimate for each metric from all the rarefaction samples for a time series.
SES estimates can be interpreted as how much of the functional characteristic (richness, evenness, divergence) was observed beyond what was expected by chance for a community of that species richness.

@@

```{r echo=FALSE}
read_csv(here::here("figures/man_model_table.csv")) %>%
  select(metric, everything()) %>%
  mutate(metric = ifelse(metric == "Jaccard_base", "Jaccard", metric),
         group = case_when(
           group == "rarefyID:study_id" ~ "timeseries:study",
           group == "study_id" ~ "study",
           TRUE ~ group),
         term = case_when(
           term == "(Intercept)" ~ "Intercept",
           term == "year_scaled" ~ "Year",
           term == "sd__(Intercept)" ~ "Intercept sd",
           term == "cor__(Intercept).year_scaled" ~ "Intercept Year corr",
           term == "sd__year_scaled" ~ "Year sd",
           term == "sd__Observation" ~ "Observation sd",
           TRUE ~ term
         )) %>%
  arrange(metric) %>%
  kable(format = "latex", caption = "Model estimates and statistics for general trend models for species richness (S) and standardized metrics. Additional model estimates can be found in the supplement.") %>%
  collapse_rows(columns = 1, valign = "middle") %>%
  kable_classic() %>% 
  kable_styling(latex_options= c("scale_down", "hold_position"))
```

@@

<!-- We also need theory that links ecological process to the kinds of changes actually happening in communities. Do the shifts we see matter? Are there shifts we didn't even measure that are critical for function and resilience? -->
